
---
title: "Idaho Heat, Drought, and Fire Metrics Report (By Question)"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
geometry:
  - margin=1in
header-includes:
  - \usepackage{graphicx}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.6,
  dpi = 300,
  fig.align = "center",
  out.width = "100%"
)
```

# Overview

This report extends the prior **fire-only** workflow to an Idaho-only, question-framed climate analysis for:

- **Heat**
- **Drought**
- **Fire**

The analysis is restricted to the five Idaho towns requested for this stage:

- **Bovill**
- **Deary**
- **Elk River**
- **Juliaetta**
- **Kendrick**

using both the town polygon and the **10-mile buffer** where available.

The pipeline now does four things in one run:

1. Reads Idaho **GridMET** daily data for the five towns.
2. Builds annual water-year metrics for **heat, drought, and fire**.
3. Runs the Christine-style **Delta / ARIMA** trend diagnostic for every mapped metric.
4. If the survey file is present, computes **PDI** outputs using the climate-comparable perception questions only.

This report uses the updated source script:

- `src/extract_fire_metrics2_heat_drought_idaho.R`

```{r source-pipeline, include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(stringr)

project_root <- getwd()
script_path <- file.path(project_root, "src", "extract_fire_metrics2_heat_drought_idaho.R")
out_dir     <- file.path(project_root, "outputs", "idaho_questions")
plot_dir    <- file.path(project_root, "plots", "idaho_questions")

stopifnot(file.exists(script_path))
source(script_path)
```

# Scope and assumptions

This version reflects the decisions made for this Idaho extension:

- **Towns included**: Deary, Bovill, Kendrick, Juliaetta, Elk River
- **Town excluded**: Troy
- **Heat metrics**: standard temperature metrics only
- **Question scope**: only the climate-comparable perception questions already mapped in the metrics spreadsheet
- **Timing / direction handling**:
  - the **instrumentation side** is sign-adjusted so that larger transformed values indicate stronger climate stress
  - survey items are currently treated as **higher = more / earlier / longer / drier / hotter / more fire**, matching the working assumption already used in the fire workflow

# Data and derived objects

```{r objects-summary}
obj_summary <- tibble(
  object = c("annual_metrics", "annual_long", "arima_outputs", "survey_q_map", "question_metric_map"),
  rows = c(nrow(annual_metrics), nrow(annual_long), nrow(arima_outputs), nrow(survey_q_map), nrow(question_metric_map)),
  cols = c(ncol(annual_metrics), ncol(annual_long), ncol(arima_outputs), ncol(survey_q_map), ncol(question_metric_map))
)

knitr::kable(obj_summary, caption = "Core objects generated by the updated Idaho pipeline")
```

# Question map

```{r question-map}
question_table <- survey_q_map %>%
  arrange(stressor, question_id) %>%
  select(stressor, question_id, survey_col, question_text)

knitr::kable(question_table, caption = "Climate-comparable survey questions used in this report")
```

# Metric map

```{r metric-map}
metric_table <- question_metric_map %>%
  arrange(stressor, question_id, metric) %>%
  select(stressor, question_id, survey_col, metric, metric_sign)

knitr::kable(metric_table, caption = "Question-to-metric map used for ARIMA and PDI")
```

# Results by stressor and question

```{r question-results, results='asis'}
library(glue)

show_question_block <- function(qid) {
  q_meta <- survey_q_map %>% filter(question_id == qid) %>% slice(1)
  q_metrics <- question_metric_map %>% filter(question_id == qid) %>% distinct(metric, metric_sign)
  q_arima <- arima_outputs %>% filter(question_id == qid)

  cat(glue("## {q_meta$question_id}: {q_meta$question_text}\n\n"))
  cat(glue("**Stressor:** {q_meta$stressor}  \n"))
  cat(glue("**Survey column:** `{q_meta$survey_col}`  \n\n"))

  cat("### Metrics used\n\n")
  print(knitr::kable(q_metrics, col.names = c("Metric", "Metric sign")))

  cat("\n\n### ARIMA summary by town / metric / buffer\n\n")
  q_tab <- q_arima %>%
    select(town, buffer, metric, sen_slope_per_decade, qcat, delta_late_minus_early) %>%
    arrange(buffer, town, metric)
  print(knitr::kable(q_tab, digits = 3))

  cat("\n\n### Mean ARIMA bucket by town (buffer = none)\n\n")
  p1 <- q_arima %>%
    filter(buffer == "none") %>%
    mutate(qcat_num = suppressWarnings(as.integer(str_remove(qcat, "^Q")))) %>%
    group_by(town) %>%
    summarise(mean_qcat = mean(qcat_num, na.rm = TRUE), .groups = "drop") %>%
    ggplot(aes(x = reorder(town, mean_qcat), y = mean_qcat)) +
    geom_col() +
    coord_flip() +
    theme_bw(base_size = 11) +
    labs(x = NULL, y = "Mean Qcat", title = paste0(qid, ": Mean ARIMA bucket by town"))

  print(p1)

  first_metric <- q_metrics$metric[1]
  metric_path <- file.path(plot_dir, paste0(first_metric, ".png"))
  if (file.exists(metric_path)) {
    cat("\n\n### Example time-series plot\n\n")
    cat(knitr::include_graphics(metric_path))
    cat("\n\n")
  }

  cat("\n\n")
}

for (st in c("Heat", "Drought", "Fire")) {
  cat(glue("# {st}\n\n"))
  qids <- survey_q_map %>% filter(stressor == st) %>% pull(question_id)
  for (qid in qids) show_question_block(qid)
}
```

# Summary counts by question and Q category

```{r qcat-summary}
qcat_summary <- arima_outputs %>%
  count(stressor, question_id, buffer, qcat, name = "n") %>%
  arrange(stressor, question_id, buffer, qcat)

knitr::kable(qcat_summary, caption = "Count of metric-level ARIMA categories by question")
```

# PDI outputs

```{r pdi-section, results='asis'}
if (exists("pdi_town")) {
  cat("## Town-level PDI\n\n")

  pdi_tab <- pdi_town %>%
    select(instr_metric, stressor, question_id, town, perception_mean, instr_value, PDI) %>%
    arrange(instr_metric, stressor, question_id, town)

  print(knitr::kable(pdi_tab, digits = 3, caption = "Town-level PDI outputs"))

  p_heat <- pdi_town %>%
    filter(instr_metric == "Qcat") %>%
    ggplot(aes(x = question_id, y = town, fill = PDI)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.2f", PDI)), size = 3) +
    facet_wrap(~stressor, scales = "free_x") +
    theme_bw(base_size = 11) +
    labs(title = "Town × Question PDI (Qcat instrumentation)", x = "Question", y = "Town")

  print(p_heat)

  p_rank <- pdi_town %>%
    filter(instr_metric == "Qcat") %>%
    group_by(town) %>%
    summarise(mean_PDI = mean(PDI, na.rm = TRUE), .groups = "drop") %>%
    mutate(town = reorder(town, mean_PDI)) %>%
    ggplot(aes(x = town, y = mean_PDI)) +
    geom_col() +
    coord_flip() +
    theme_bw(base_size = 11) +
    labs(title = "Mean PDI across all mapped questions", x = NULL, y = "Mean PDI")

  print(p_rank)
} else {
  cat("Survey file not found at knit time, so PDI tables/plots were skipped.\n")
}
```

# Output files written by the pipeline

```{r output-files}
out_files <- tibble(file = sort(list.files(out_dir)))
knitr::kable(out_files, caption = "Files written to outputs/idaho_questions")
```

# Notes

- This report intentionally stays focused on the **climate-comparable perception questions** only.
- The heat extension currently excludes **UTCI-based** metrics.
- Timing metrics that imply **earlier onset** but are numerically smaller were sign-adjusted on the instrumentation side before ARIMA and PDI.
- The survey direction assumptions for timing questions are still the most important place to revisit if you later want to tighten the inferential setup.
