---
title: "Idaho Climate Metrics Report (By Question)"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
geometry:
  - margin=1in
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{graphicx}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6.5,
  fig.height = 4.2,
  dpi = 300,
  fig.align = 'center',
  out.width = '100%'
)
```

# Overview

This report produces a **question-framed** summary of **heat, drought, and fire** metrics for the five Idaho towns:

- Deary
- Bovill
- Kendrick
- Juliaetta
- Elk River

The structure follows the original **fire-only report** as closely as possible, while extending the analysis to the climate-comparable survey questions for **heat, drought, and fire**. The emphasis here is on **question-by-question plots**, with each metric shown **by town and buffer**, rather than large summary tables at the start of each section.

This RMarkdown is designed so that knitting the PDF runs the working Idaho script once, then reads the saved outputs and embeds the saved figures.

# Data sources and conventions

## Data source

- **GridMET** daily climate time series for each town and its 10-mile buffer.
- Expected input files follow the existing project naming convention:
  - `Town_gridmet.csv`
  - `Town_10mibuffer_gridmet.csv`

## Calendar and water-year conventions

- **Water year**: Oct 1 - Sep 30, labeled by the ending year.
- **wyday**: water-year day where Oct 1 = 1.
- **season4**: DJF / MAM / JJA / SON.
- **baseline climatology period**: 1996-2025.

```{r source-pipeline, include=FALSE}
library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(tidyr)
library(lubridate)
library(zoo)
library(ggplot2)
library(trend)
library(forecast)
library(knitr)
library(tibble)

project_root <- getwd()

script_candidates <- c(
  file.path(project_root, "src", "extract_idaho_heat_drought_fire_metrics.R"),
  file.path(project_root, "src", "extract_idaho_heat_drought_fire_metrics_v2.R"),
  file.path(project_root, "src", "extract_fire_metrics2_heat_drought_idaho.R")
)

script_path <- script_candidates[file.exists(script_candidates)][1]
stopifnot(!is.na(script_path), file.exists(script_path))

plot_dir <- file.path(project_root, "plots")
out_dir  <- file.path(project_root, "outputs")

source(script_path)

question_outputs_path <- file.path(out_dir, "idaho_question_outputs.csv")
pdi_summary_path      <- file.path(out_dir, "idaho_pdi_summary.csv")
pdi_person_path       <- file.path(out_dir, "idaho_pdi_person.csv")

stopifnot(file.exists(question_outputs_path))
stopifnot(file.exists(pdi_summary_path))
stopifnot(file.exists(pdi_person_path))

question_outputs <- readr::read_csv(question_outputs_path, show_col_types = FALSE)
pdi_summary      <- readr::read_csv(pdi_summary_path, show_col_types = FALSE)
pdi_person       <- readr::read_csv(pdi_person_path, show_col_types = FALSE)
```

# Metric construction and question mapping

## Heat questions

- **H1**: Overall heat magnitude
- **H2**: Summer heat magnitude
- **H3**: Heat frequency
- **H4**: Spring heat onset timing
- **H5**: Fall cooling timing
- **H6**: Heat duration / persistence

## Drought questions

- **D1**: Overall wetter/drier magnitude
- **D2**: Winter wetter/drier magnitude
- **D3**: Winter dryness frequency
- **D4**: Precipitation timing
- **D5**: Dry-duration persistence

## Fire questions

- **F1**: Overall frequency of high fire-weather days
- **F2**: Summer frequency of high fire-weather days
- **F3**: Fire-weather magnitude / intensity proxy
- **F4**: Fire-season start timing
- **F5**: Fire-season duration / persistence

*(Question-to-metric mapping table suppressed to keep the report concise.)*

# Results: Question-framed plots

The script saves question-specific plots to `plots/`. Below, each question is shown in the same spirit as the original fire report: short interpretation text followed by the saved **by-town** time-series figures.

```{r helpers, include=FALSE}
question_meta <- tribble(
  ~question, ~title, ~survey_question, ~interpretation, ~metrics,
  "H1", "Overall heat magnitude", "Overall, has it gotten hotter or cooler?", "Higher values indicate stronger overall warming relative to baseline conditions.", list(c("meananntemp", "meananomanntemp", "meanmaxanntemp", "meanmaxanomanntemp", "meanminanntemp", "meanminanomanntemp")),
  "H2", "Summer heat magnitude", "During the summer months, has it gotten hotter or cooler?", "Higher values indicate hotter summer conditions, especially during JJA.", list(c("maxmaxanomjjatemp", "maxmaxjjatemp", "maxminanomjjatemp", "maxminjjatemp", "meananomjjatemp", "meanjjatemp")),
  "H3", "Heat frequency", "If summer has changed, how frequently is it hotter?", "Higher values indicate more frequent hot-day conditions.", list(c("frostdays", "summerdays", "tmaxgt90thdays", "tmingt90thdays")),
  "H4", "Spring heat onset timing", "When does it start getting hotter in the spring?", "Smaller day-of-year values indicate earlier spring onset.", list(c("gsstartdoy")),
  "H5", "Fall cooling timing", "When does it start getting cooler in the fall?", "Larger end-day values indicate later persistence of warm-season conditions.", list(c("gsenddoy")),
  "H6", "Heat duration / persistence", "Do heat conditions last shorter or longer than you remember?", "Higher values indicate longer hot periods or longer warm-season persistence.", list(c("durtmaxgt90th", "gslength")),
  "D1", "Overall wetter/drier magnitude", "Overall, has it gotten wetter or drier?", "Lower precipitation and larger moisture deficits indicate stronger drought stress.", list(c("annualanometo", "annualanomprecip", "annualanompwd", "annualeto", "annualprecip", "annualpwd")),
  "D2", "Winter wetter/drier magnitude", "During the winter months, has it become wetter or drier?", "Focuses on cool-season precipitation conditions.", list(c("coolseasonanomprecip", "coolseasonprecip")),
  "D3", "Winter dryness frequency", "If winter has changed, how frequently is it drier?", "Higher values indicate more frequent winter dry-day conditions.", list(c("coolseasondrydays1mm")),
  "D4", "Precipitation timing", "Has there been a change in when rainfall/snowfall starts during the year?", "Timing metrics describe when precipitation tends to begin or concentrate within the water year.", list(c("preciponsetwydoy", "precipCOMwydoy")),
  "D5", "Dry-duration persistence", "Do drier conditions last shorter or longer than you remember?", "Higher values indicate longer dry spells or more persistent drought conditions.", list(c("maxdrydur", "maxcoolseasondrydur")),
  "F1", "Overall frequency of high fire-weather days", "Overall, has fire become less common or more common?", "Higher values indicate more days in the year with unusually elevated fire-weather conditions.", list(c("firehighdays_erc90", "firehighdays_bi90", "firehighdays_erc95", "firehighdays_bi95")),
  "F2", "Summer frequency of high fire-weather days", "During the summer months, has fire occurred more often or less often?", "Focuses on the summer portion of the year when fire-weather conditions are often most consequential.", list(c("firesummerhighdays_erc90", "firesummerhighdays_bi90")),
  "F3", "Fire-weather magnitude / intensity proxy", "Have the areas affected by wildfires gotten smaller or larger?", "These metrics summarize how extreme fire-weather conditions become within a year.", list(c("firepeak_erc", "firepeak_bi", "firep95_erc", "fireseverity_erc90", "fireseverity_bi90")),
  "F4", "Fire-season start timing", "Has there been a change in when the fire season starts during the year?", "Smaller water-year day values indicate earlier seasonal onset.", list(c("firestart_wyday_erc", "firestart_wyday_bi", "firestart_any_erc90", "firestart_any_bi90", "firestart_k2_erc90", "firestart_k3_erc90", "firestart_k3_bi90")),
  "F5", "Fire-season duration / persistence", "Do the fire conditions last shorter or longer than you remember?", "Higher values indicate longer or more persistent periods of elevated fire-weather conditions.", list(c("fireseasonlength_erc", "fireseasonlength_bi", "firemaxrun_erc", "firemaxrun_bi"))
)

include_plot <- function(fname) {
  f <- file.path(plot_dir, fname)
  if (!file.exists(f)) {
    cat("(Plot not found:", fname, ")\n")
    return(invisible(NULL))
  }
  knitr::include_graphics(f)
}

show_question_plots <- function(qcode) {
  row <- question_meta %>% filter(question == qcode)
  mets <- row$metrics[[1]]
  for (m in mets) {
    include_plot(paste0(qcode, "_timeseries_", m, ".png"))
  }
}
```

## H1: Overall heat magnitude

**Survey question:** “Overall, has it gotten hotter or cooler?”

**Interpretation:** Higher values indicate stronger overall warming relative to baseline conditions.

**Metrics used:** `meananntemp`, `meananomanntemp`, `meanmaxanntemp`, `meanmaxanomanntemp`, `meanminanntemp`, `meanminanomanntemp`

```{r h1-plots}
show_question_plots("H1")
```

## H2: Summer heat magnitude

**Survey question:** “During the summer months, has it gotten hotter or cooler?”

**Interpretation:** Higher values indicate hotter summer conditions, especially during JJA.

**Metrics used:** `maxmaxanomjjatemp`, `maxmaxjjatemp`, `maxminanomjjatemp`, `maxminjjatemp`, `meananomjjatemp`, `meanjjatemp`

```{r h2-plots}
show_question_plots("H2")
```

## H3: Heat frequency

**Survey question:** “If summer has changed, how frequently is it hotter?”

**Interpretation:** Higher values indicate more frequent hot-day conditions.

**Metrics used:** `frostdays`, `summerdays`, `tmaxgt90thdays`, `tmingt90thdays`

```{r h3-plots}
show_question_plots("H3")
```

## H4: Spring heat onset timing

**Survey question:** “When does it start getting hotter in the spring?”

**Interpretation:** Smaller day-of-year values indicate earlier spring onset.

**Metrics used:** `gsstartdoy`

```{r h4-plots}
show_question_plots("H4")
```

## H5: Fall cooling timing

**Survey question:** “When does it start getting cooler in the fall?”

**Interpretation:** Larger day-of-year values indicate later seasonal cooling.

**Metrics used:** `gsenddoy`

```{r h5-plots}
show_question_plots("H5")
```

## H6: Heat duration / persistence

**Survey question:** “Do heat conditions last shorter or longer than you remember?”

**Interpretation:** Higher values indicate longer hot periods or longer warm-season persistence.

**Metrics used:** `durtmaxgt90th`, `gslength`

```{r h6-plots}
show_question_plots("H6")
```

## D1: Overall wetter/drier magnitude

**Survey question:** “Overall, has it gotten wetter or drier?”

**Interpretation:** Lower precipitation and larger moisture deficits indicate stronger drought stress.

**Metrics used:** `annualanometo`, `annualanomprecip`, `annualanompwd`, `annualeto`, `annualprecip`, `annualpwd`

```{r d1-plots}
show_question_plots("D1")
```

## D2: Winter wetter/drier magnitude

**Survey question:** “During the winter months, has it become wetter or drier?”

**Interpretation:** Focuses on cool-season precipitation conditions.

**Metrics used:** `coolseasonanomprecip`, `coolseasonprecip`

```{r d2-plots}
show_question_plots("D2")
```

## D3: Winter dryness frequency

**Survey question:** “If winter has changed, how frequently is it drier?”

**Interpretation:** Higher values indicate more frequent winter dry-day conditions.

**Metrics used:** `coolseasondrydays1mm`

```{r d3-plots}
show_question_plots("D3")
```

## D4: Precipitation timing

**Survey question:** “Has there been a change in when rainfall/snowfall starts during the year?”

**Interpretation:** Timing metrics describe when precipitation tends to begin or concentrate within the water year.

**Metrics used:** `preciponsetwydoy`, `precipCOMwydoy`

```{r d4-plots}
show_question_plots("D4")
```

## D5: Dry-duration persistence

**Survey question:** “Do drier conditions last shorter or longer than you remember?”

**Interpretation:** Higher values indicate longer dry spells or more persistent drought conditions.

**Metrics used:** `maxdrydur`, `maxcoolseasondrydur`

```{r d5-plots}
show_question_plots("D5")
```

## F1: Overall frequency of high fire-weather days

**Survey question:** “Overall, has fire become less common or more common?”

**Interpretation:** Higher values indicate more days in the year with unusually elevated fire-weather conditions.

**Metrics used:** `firehighdays_erc90`, `firehighdays_bi90`, `firehighdays_erc95`, `firehighdays_bi95`

```{r f1-plots}
show_question_plots("F1")
```

## F2: Summer frequency of high fire-weather days

**Survey question:** “During the summer months, has fire occurred more often or less often?”

**Interpretation:** Focuses on the summer portion of the year when fire-weather conditions are often most consequential.

**Metrics used:** `firesummerhighdays_erc90`, `firesummerhighdays_bi90`

```{r f2-plots}
show_question_plots("F2")
```

## F3: Fire-weather magnitude / intensity proxy

**Survey question:** “Have the areas affected by wildfires gotten smaller or larger?”

**Interpretation:** These metrics summarize how extreme fire-weather conditions become within a year.

**Metrics used:** `firepeak_erc`, `firepeak_bi`, `firep95_erc`, `fireseverity_erc90`, `fireseverity_bi90`

```{r f3-plots}
show_question_plots("F3")
```

## F4: Fire-season start timing

**Survey question:** “Has there been a change in when the fire season starts during the year?”

**Interpretation:** Smaller water-year day values indicate earlier seasonal onset.

**Metrics used:** `firestart_wyday_erc`, `firestart_wyday_bi`, `firestart_any_erc90`, `firestart_any_bi90`, `firestart_k2_erc90`, `firestart_k3_erc90`, `firestart_k3_bi90`

```{r f4-plots}
show_question_plots("F4")
```

## F5: Fire-season duration / persistence

**Survey question:** “Do the fire conditions last shorter or longer than you remember?”

**Interpretation:** Higher values indicate longer or more persistent periods of elevated fire-weather conditions.

**Metrics used:** `fireseasonlength_erc`, `fireseasonlength_bi`, `firemaxrun_erc`, `firemaxrun_bi`

```{r f5-plots}
show_question_plots("F5")
```

# Results: Delta / ARIMA outputs

## What the Delta / ARIMA block is doing

For each metric (town × buffer):

1. Compute baseline anomalies relative to the mean over 1996-2025.
2. Compute Sen’s slope (per decade) for the observed series.
3. Fit an ARIMA model with no drift.
4. Simulate many synthetic series from the fitted ARIMA model.
5. Place the observed slope into Q1-Q7 bins based on the simulated distribution.

The resulting `qcat` summarizes how unusual the observed trend is relative to the fitted stationary ARIMA process.

## ARIMA results: visual summaries

```{r arima-prep, include=FALSE}
q_order <- c(paste0("H", 1:6), paste0("D", 1:5), paste0("F", 1:5))
question_outputs <- question_outputs %>%
  mutate(
    question = factor(question, levels = q_order),
    qcat = factor(qcat, levels = paste0("Q", 1:7), ordered = TRUE),
    qcat_num = as.integer(qcat),
    unit = paste(town, buffer, sep = " | ")
  )
```

```{r arima-qcat-heatmap}
ggplot(question_outputs, aes(x = question, y = unit, fill = qcat_num)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_continuous(breaks = 1:7, limits = c(1, 7)) +
  labs(
    title = "ARIMA Delta diagnostics: Q1-Q7 category by question and location",
    x = "Question",
    y = "Town | Buffer",
    fill = "Qcat (1-7)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.y = element_text(size = 7),
    panel.grid = element_blank()
  )
```

```{r arima-qcat-counts}
question_outputs %>%
  count(question, qcat) %>%
  ggplot(aes(x = question, y = n, fill = qcat)) +
  geom_col(position = "stack") +
  labs(
    title = "Counts of ARIMA Qcat categories by question",
    x = "Question",
    y = "Count",
    fill = "Qcat"
  ) +
  theme_minimal(base_size = 11) +
  scale_fill_discrete(drop = FALSE)
```

# PDI (Perception Δ Instrumentation)

This section summarizes the tenure-based PDI outputs from the working Idaho script.

```{r pdi-summary-table}
pdi_summary %>%
  select(question, .town, n, mean_resp, mean_instr_delta_raw, mean_climate_delta_z, mean_pdi, sd_pdi) %>%
  arrange(question, .town) %>%
  knitr::kable(
    caption = "Town-level PDI summary by question.",
    digits = 3,
    align = c("l", "l", "r", "r", "r", "r", "r", "r")
  )
```

```{r pdi-plot}
pdi_summary %>%
  mutate(question = factor(question, levels = q_order)) %>%
  ggplot(aes(x = .town, y = mean_pdi)) +
  geom_col() +
  facet_wrap(~ question, scales = "free_y") +
  labs(
    title = "Mean PDI by town and question",
    x = "Town",
    y = "Mean PDI"
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Notes / interpretation guidance

- Frequency metrics generally increase when hot, dry, or high fire-weather days become more common.
- Timing metrics require directional interpretation: earlier onset often corresponds to smaller day values.
- Magnitude metrics summarize intensity or severity, not direct impacts such as burned area.
- Qcat values summarize how unusual an observed trend is relative to simulated ARIMA variability.
- PDI values indicate the degree of mismatch between reported perceptions and climate-derived instrumentation.
