---
title: "Idaho Climate Metrics Report (By Question)"
output:
  pdf_document:
    toc: true
    number_sections: true
geometry: margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}
  - \usepackage{needspace}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)
```

```{r source-pipeline, include=FALSE}
library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(tidyr)
library(lubridate)
library(zoo)
library(ggplot2)
library(trend)
library(forecast)
library(knitr)

project_root <- getwd()

script_candidates <- file.path(project_root, "src", "extract_idaho_heat_drought_fire_metrics_v2.R")

script_path <- script_candidates[file.exists(script_candidates)][1]
stopifnot(!is.na(script_path), file.exists(script_path))

plot_dir <- file.path(project_root, "plots")
out_dir  <- file.path(project_root, "outputs")

source(script_path)
```

```{r load-outputs, include=FALSE}
question_outputs <- readr::read_csv(file.path(out_dir, "idaho_question_outputs.csv"), show_col_types = FALSE)
pdi_summary <- readr::read_csv(file.path(out_dir, "idaho_pdi_summary.csv"), show_col_types = FALSE)
pdi_person <- readr::read_csv(file.path(out_dir, "idaho_pdi_person.csv"), show_col_types = FALSE)
annual_long <- readr::read_csv(file.path(out_dir, "idaho_annual_metrics_long.csv"), show_col_types = FALSE)

question_order <- c(paste0("H", 1:6), paste0("D", 1:5), paste0("F", 1:5))
question_outputs <- question_outputs %>% mutate(question = factor(question, levels = question_order))
pdi_summary <- pdi_summary %>% mutate(question = factor(question, levels = question_order))

question_meta <- tibble::tribble(
  ~section, ~question, ~title, ~survey_question, ~interpretation,
  "Heat", "H1", "Overall heat magnitude", "Overall, has it gotten hotter or cooler?", "Higher values indicate stronger overall warming relative to baseline conditions.",
  "Heat", "H2", "Summer heat magnitude", "During the summer months, has it gotten hotter or cooler?", "Higher values indicate hotter summer conditions, especially in JJA.",
  "Heat", "H3", "Heat frequency", "If summer has changed, how frequently is it hotter?", "Higher values indicate more frequent hot days or warm-threshold exceedances.",
  "Heat", "H4", "Spring heat onset timing", "When does it start getting hotter in the spring?", "Smaller timing values generally indicate earlier onset of spring warmth.",
  "Heat", "H5", "Fall cooling timing", "When does it start getting cooler in the fall?", "Larger timing values generally indicate later seasonal cooling.",
  "Heat", "H6", "Heat duration / persistence", "Do heat conditions last shorter or longer than you remember?", "Higher values indicate longer hot periods or longer warm seasons.",
  "Drought", "D1", "Overall wetter/drier magnitude", "Overall, has it gotten wetter or drier?", "Lower precipitation or larger moisture deficits indicate stronger drought stress.",
  "Drought", "D2", "Winter wetter/drier magnitude", "During the winter months, has it become wetter or drier?", "Lower cool-season precipitation or stronger deficits indicate drier winters.",
  "Drought", "D3", "Winter dryness frequency", "If winter has changed, how frequently is it drier?", "Higher values indicate more frequent dry winter days or low-precipitation conditions.",
  "Drought", "D4", "Precipitation timing", "Has there been a change in when rainfall/snowfall starts during the year?", "Timing metrics describe when precipitation arrives within the water year.",
  "Drought", "D5", "Dry-duration persistence", "Do drier conditions last shorter or longer than you remember?", "Higher values indicate longer uninterrupted dry periods.",
  "Fire", "F1", "Overall frequency of high fire-weather days", "Overall, has fire become less common or more common?", "Higher values indicate more days with unusually elevated fire-weather conditions.",
  "Fire", "F2", "Summer frequency of high fire-weather days", "During summer, has fire occurred more often or less often?", "Higher values indicate more frequent high fire-weather during JJA.",
  "Fire", "F3", "Fire-weather magnitude / intensity proxy", "Have the areas affected by wildfires gotten smaller or larger?", "Higher values indicate more intense fire-weather conditions, not direct burned area.",
  "Fire", "F4", "Fire-season start timing", "Has there been a change in when the fire season starts during the year?", "Smaller timing values generally indicate earlier onset of the fire-weather season.",
  "Fire", "F5", "Fire-season duration / persistence", "Do fire conditions last shorter or longer than you remember?", "Higher values indicate longer runs or longer seasonal duration of elevated fire-weather."
)

render_plot_set <- function(q) {
  mets <- question_outputs %>%
    filter(question == q) %>%
    distinct(metric) %>%
    arrange(metric) %>%
    pull(metric)

  for (m in mets) {
    p <- file.path(plot_dir, paste0(as.character(q), "_timeseries_", m, ".png"))
    if (file.exists(p)) {
      cat("\n\n")
      cat("\\Needspace{0.45\textheight}\
")
      cat("\\begin{figure}[H]\
")
      cat("\\centering\
")
      cat("\\textbf{", m, "}\par\medskip\
", sep = "")
      cat("\\includegraphics[width=0.98\textwidth]{", p, "}\
", sep = "")
      cat("\\end{figure}\
")
      cat("\\FloatBarrier\
")
      cat("\n\n")
    } else {
      cat("\n\n_Missing plot:_ `", p, "`\n\n", sep = "")
    }
  }
}
```

# Overview

This report produces a question-framed summary of heat, drought, and fire metrics for the five Idaho towns: Deary, Bovill, Kendrick, Juliaetta, and Elk River. The workflow follows the same general style as the earlier fire-only report, but extends the analysis to the climate-comparable heat and drought survey questions as well.

# Data sources and conventions

## Data source

- GridMET daily climate time series for each town and its 10-mile buffer.
- Input files follow the existing project naming convention used in the working script.

## Calendar and water-year conventions

- Water year: Oct 1 to Sep 30, labeled by the ending year.
- `wyday`: water-year day where Oct 1 = 1.
- `season4`: DJF / MAM / JJA / SON.
- Baseline climatology period: 1996 to 2025.

# Metric construction and question mapping

## Heat questions

- H1: Overall heat magnitude
- H2: Summer heat magnitude
- H3: Heat frequency
- H4: Spring heat onset timing
- H5: Fall cooling timing
- H6: Heat duration / persistence

## Drought questions

- D1: Overall wetter/drier magnitude
- D2: Winter wetter/drier magnitude
- D3: Winter dryness frequency
- D4: Precipitation timing
- D5: Dry-duration persistence

## Fire questions

- F1: Overall frequency of high fire-weather days
- F2: Summer frequency of high fire-weather days
- F3: Fire-weather magnitude / intensity proxy
- F4: Fire-season start timing
- F5: Fire-season duration / persistence

# Results: Question-framed plots

# Heat

```{r heat-sections, results='asis'}
for (q in question_meta$question[question_meta$section == "Heat"]) {
  meta <- question_meta %>% filter(question == q)
  cat("## ", q, ": ", meta$title[[1]], "\n\n", sep = "")
  cat("Survey question: \"", meta$survey_question[[1]], "\"\n\n", sep = "")
  cat("Interpretation: ", meta$interpretation[[1]], "\n\n", sep = "")
  cat("Metrics used (variable names):\n\n")
  mets <- question_outputs %>% filter(question == q) %>% distinct(metric) %>% arrange(metric) %>% pull(metric)
  for (m in mets) cat("- `", m, "`\n", sep = "")
  cat("\n")
  render_plot_set(q)
}
```

# Drought

```{r drought-sections, results='asis'}
for (q in question_meta$question[question_meta$section == "Drought"]) {
  meta <- question_meta %>% filter(question == q)
  cat("## ", q, ": ", meta$title[[1]], "\n\n", sep = "")
  cat("Survey question: \"", meta$survey_question[[1]], "\"\n\n", sep = "")
  cat("Interpretation: ", meta$interpretation[[1]], "\n\n", sep = "")
  cat("Metrics used (variable names):\n\n")
  mets <- question_outputs %>% filter(question == q) %>% distinct(metric) %>% arrange(metric) %>% pull(metric)
  for (m in mets) cat("- `", m, "`\n", sep = "")
  cat("\n")
  render_plot_set(q)
}
```

# Fire

```{r fire-sections, results='asis'}
for (q in question_meta$question[question_meta$section == "Fire"]) {
  meta <- question_meta %>% filter(question == q)
  cat("## ", q, ": ", meta$title[[1]], "\n\n", sep = "")
  cat("Survey question: \"", meta$survey_question[[1]], "\"\n\n", sep = "")
  cat("Interpretation: ", meta$interpretation[[1]], "\n\n", sep = "")
  cat("Metrics used (variable names):\n\n")
  mets <- question_outputs %>% filter(question == q) %>% distinct(metric) %>% arrange(metric) %>% pull(metric)
  for (m in mets) cat("- `", m, "`\n", sep = "")
  cat("\n")
  render_plot_set(q)
}
```

# Delta / ARIMA outputs

## What the Delta / ARIMA block is doing

For each town, buffer, and annual metric, the workflow estimates a trend summary and compares the observed Sen slope to a simulated ARIMA reference distribution. The `qcat` values indicate where the observed slope falls relative to that simulated distribution.

```{r arima-visual-1, fig.height=6, fig.width=9}
question_outputs %>%
  mutate(label = paste(town, buffer, sep = " | ")) %>%
  ggplot(aes(x = question, y = as.numeric(str_remove(qcat, "Q")), fill = buffer)) +
  geom_col(position = position_dodge(width = 0.8)) +
  facet_wrap(~ town, ncol = 2) +
  scale_y_continuous(breaks = 1:7) +
  labs(title = "ARIMA Delta diagnostics: Qcat by question and town/buffer",
       x = "Question", y = "Qcat (1-7)")
```

```{r arima-visual-2, fig.height=5, fig.width=8}
question_outputs %>%
  count(question, qcat) %>%
  ggplot(aes(x = question, y = n, fill = qcat)) +
  geom_col() +
  labs(title = "Counts of ARIMA Qcat categories by question",
       x = "Question", y = "Count")
```

# PDI (Perception Delta Instrumentation)

This section summarizes the mismatch between survey perceptions and tenure-linked climate instrumentation.

```{r pdi-summary-table}
knitr::kable(
  pdi_summary %>% arrange(question, .town),
  caption = "Town-level PDI summary by question"
)
```

```{r pdi-mean-plot, fig.height=7, fig.width=9}
pdi_summary %>%
  ggplot(aes(x = .town, y = mean_pdi)) +
  geom_col() +
  facet_wrap(~ question, scales = "free_y") +
  labs(title = "Mean PDI by town and question", x = "Town", y = "Mean PDI")
```

# Notes / interpretation guidance

- Frequency metrics generally increase when hot, dry, or high fire-weather days become more common.
- Timing metrics require directional interpretation: earlier onset often corresponds to smaller day values.
- Magnitude metrics summarize intensity or severity, not direct impacts such as burned area.
- `qcat` values summarize how unusual an observed trend is relative to simulated ARIMA variability.
- PDI values indicate the degree of mismatch between reported perceptions and climate-derived instrumentation.
