---
title: "Fire Metrics Report (By Question)"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
geometry:
  - margin=1in
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{graphicx}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6.5,
  fig.height = 4.2,
  dpi = 300,
  fig.align = 'center',
  out.width = '100%'
)
```

# Overview

This report produces a **question-framed** summary of *fire-weather* metrics derived from **GridMET** daily time series (per town and buffer). The workflow mirrors Christine’s approach used for heat/drought metrics:

1. **Parse daily dates** from `system.index` and create water-year variables.
2. Build a **daily climatology** (1996-2025) and use it to define *relative* “high fire-weather days” (e.g., above the daily 90th percentile).
3. Aggregate daily flags into **annual (water-year) metrics**.
4. Organize the metrics into **survey-question buckets** (Q1-Q5) with plots and summary tables.
5. At the end, run the **Christine-style Delta/ARIMA diagnostic** to produce Q1-Q7 quantile categories for trend strength.

This RMarkdown is designed so that **knitting the PDF runs the full pipeline once** by sourcing the up-to-date script (`src/extract_fire_metrics2.R`), and then **renders plots/tables in the report** from the generated objects and saved outputs.

# Data sources and fire-weather variables

## Data source

- **GridMET**: gridded daily meteorology (used here for fire-weather indices and supporting variables).
- Input files are expected under `data/ID_data/ID/` and follow the naming convention:
  - `Town_gridmet.csv`
  - `Town_10mibuffer_gridmet.csv`
  - (Other sources can exist, but fire metrics in this report are computed from **gridmet** only.)

## Fire-weather indices used

These are **fire danger / fire-weather** indicators, not direct measurements of burned area:

- **BI (Burning Index)**: composite index related to potential fire behavior (higher BI - more intense potential fire behavior).
- **ERC (Energy Release Component)**: proxy for available energy in the flaming front; responds strongly to fuel dryness (higher ERC - more potential energy release).
- **FM100 / FM1000**: 100-hr and 1000-hr dead fuel moisture proxies (lower - drier fuels).

In this project, BI and ERC are the primary drivers for defining “high fire-weather days” using **relative exceedance** above the daily climatology percentiles.

## Calendar and water-year conventions

- **Water year**: Oct 1 - Sep 30, labeled by the ending year (e.g., Oct 2019-Sep 2020 is **WY2020**).
- **wyday**: “water-year day” where Oct 1 = 1, and Sep 30 = 365/366.
- **season4**: DJF / MAM / JJA / SON.

```{r fire-pipeline, include=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(tidyr)
library(lubridate)
library(zoo)
library(ggplot2)
library(trend)
library(forecast)
library(knitr)

# Project root: this Rmd lives at repo root; script is in /src
project_root  <- getwd()
script_path   <- file.path(project_root, "src", "extract_fire_metrics2.R")

# Where the script writes outputs
fire_plot_dir <- file.path(project_root, "plots", "fire")
out_dir       <- file.path(project_root, "outputs", "fire_questions")

stopifnot(file.exists(script_path))

# Run full fire-metrics workflow (silent in report output)
source(script_path)
```

# Metric construction and question mapping

## How “high fire-weather days” are defined

For each calendar day (month/day), a **daily climatology threshold** is computed from **1996-2025** using the chosen quantile (typically **q90**, optionally **q95**). A day is a “high day” if:

- `BI > dailyclimo9625_bi_q0.9`  (BI high day)
- `ERC > dailyclimo9625_erc_q0.9` (ERC high day)

This makes the threshold **seasonally varying**, so “high” in spring is defined relative to typical spring conditions.

## Annual metrics (water-year summaries)

Daily flags are aggregated into water-year metrics. The main families are:

- **Frequency**: number of high days per year
- **Seasonal frequency**: number of high days in summer (JJA)
- **Intensity / magnitude proxy**: annual maxima or high percentiles
- **Timing**: first sustained run of high days (start of season)
- **Duration**: maximum consecutive run length; optional season length

## Question-to-metric mapping

These are the **Christine-style “question buckets”** used to generate CSV outputs and summary tables.

```{r show-fire-map, include=FALSE}
# Mapping table suppressed in the PDF to keep the report clean.
# If you want it back, set include=TRUE.
stopifnot(exists("fire_map"))
fire_map_df <- tibble(
  Question = rep(names(fire_map), lengths(fire_map)),
  Metric   = unlist(fire_map)
)
```

*(Question-to-metric mapping table suppressed to keep the report concise.)*

# Results: Question-framed plots

The script saves a standard set of plots to `plots/fire/`. Below we embed those images in the same Q1-Q5 structure.

```{r helper-include-plot}
include_plot <- function(fname, cap = NULL) {
  f <- file.path(fire_plot_dir, fname)
  if (!file.exists(f)) {
    cat("(Plot not found:", fname, ")\n")
    return(invisible(NULL))
  }
  knitr::include_graphics(f)
}
```

## Q1: Overall frequency of high fire-weather days

**Survey question:** “Overall, has fire become less common or more common?” (frequency)

**Fire metrics used (variable names):**
- `firehighdays_erc90` — annual count of days above the daily 90th-percentile fire threshold (BI q90 or ERC q90). More days = more frequent high fire-weather.. **Source:** GridMET ERC.
- `firehighdays_bi90` — annual count of days above the daily 90th-percentile fire threshold (BI q90 or ERC q90). More days = more frequent high fire-weather.. **Source:** GridMET BI.


Interpretation: higher values indicate more days in the year where fire-weather conditions are unusually elevated relative to the daily 1996-2025 climatology.

```{r q1-plots}
include_plot("Q1_timeseries_firehighdays_erc90.png")
include_plot("Q1_timeseries_firehighdays_bi90.png")
include_plot("Q1_distribution_firehighdays_erc90.png")
```

## Q2: Summer frequency (JJA)

**Survey question:** “During the summer months, has fire occurred more often or less often?” (summer frequency)

**Fire metrics used (variable names):**
- `firesummerhighdays_erc90` — same as above, but restricted to summer (JJA).. **Source:** GridMET ERC.
- `firesummerhighdays_bi90` — same as above, but restricted to summer (JJA).. **Source:** GridMET BI.


Interpretation: focuses on the part of the year when fire activity and impacts are often highest. This is a subset of Q1 restricted to JJA.

```{r q2-plots}
include_plot("Q2_timeseries_firesummerhighdays_erc90.png")
include_plot("Q2_timeseries_firesummerhighdays_bi90.png")
```

## Q3: Magnitude / intensity proxies (not burned area)

**Survey question:** “Have the fires been smaller or larger / less or more severe?” (magnitude/impact proxy)

**Fire metrics used (variable names):**
- `firepeak_erc` — annual peak (maximum daily) BI or ERC. Higher peaks ⇒ higher potential intensity on the worst day each year.. **Source:** GridMET ERC.
- `firepeak_bi` — annual peak (maximum daily) BI or ERC. Higher peaks ⇒ higher potential intensity on the worst day each year.. **Source:** GridMET BI.
- `firep95_erc` — annual 95th percentile of daily ERC values (tail intensity). Higher values ⇒ “typical extreme” days are more intense, not just the single worst day.. **Source:** GridMET ERC.


Interpretation: these metrics capture **how extreme** fire-weather conditions become within a year.

- `firepeak_*`: annual maximum
- `firep95_erc`: upper-tail intensity (95th percentile)
- `fireseverity_*`: exceedance above the daily q90 threshold summed across the year

```{r q3-plots}
include_plot("Q3proxy_timeseries_firepeak_erc.png")
include_plot("Q3proxy_timeseries_firep95_erc.png")
include_plot("Q3proxy_timeseries_firepeak_bi.png")
```

## Q4: Timing of the fire-weather season start

**Survey question:** “Has there been a change in when the fire season starts?” (timing)

**Fire metrics used (variable names):**
- `firestart_wyday_erc` — water-year day (wyday) of the first sustained high-fire run (based on q90 threshold and your run-length rule). Smaller wyday ⇒ earlier seasonal onset.. **Source:** GridMET ERC.
- `firestart_wyday_bi` — water-year day (wyday) of the first sustained high-fire run (based on q90 threshold and your run-length rule). Smaller wyday ⇒ earlier seasonal onset.. **Source:** GridMET BI.


Interpretation: smaller `wyday` means the season starts earlier (because Oct 1 = 1). Start is defined using a **sustained run** of high days (e.g., 3 consecutive high days), which avoids treating isolated spikes as “season onset”.

```{r q4-plots}
include_plot("Q4_timeseries_firestart_wyday_erc.png")
include_plot("Q4_timeseries_firestart_wyday_bi.png")
include_plot("Q4_periodcompare_firestart_wyday_erc.png")
```

## Q5: Duration / persistence

**Survey question:** “Do the fire conditions last shorter or longer than you remember?” (duration)

**Fire metrics used (variable names):**
- `fireseasonlength_erc` — season length (last − first high-fire day) in wyday units. Larger ⇒ longer season window with high-fire conditions.. **Source:** GridMET ERC.
- `firemaxrun_erc` — maximum consecutive run length of high-fire days in a water year. Larger ⇒ longer uninterrupted high-fire “streaks”.. **Source:** GridMET ERC.
- `firemaxrun_bi` — maximum consecutive run length of high-fire days in a water year. Larger ⇒ longer uninterrupted high-fire “streaks”.. **Source:** GridMET BI.


Interpretation: these metrics describe how long high fire-weather conditions persist.

- `firemaxrun_*`: longest consecutive run of high days in a year
- `fireseasonlength_*`: (optional) span from first to last high day

```{r q5-plots}
include_plot("Q5_timeseries_firemaxrun_erc.png")
include_plot("Q5_timeseries_firemaxrun_bi.png")
include_plot("Q5_timeseries_fireseasonlength_erc.png")
```

# Results: Christine-style Delta/ARIMA outputs (CSV tables)

## What the Delta/ARIMA block is doing

For each metric (town × buffer):

1. Compute **baseline anomalies** relative to the mean over 1996-2025.
2. Compute **Sen’s slope** (per decade) for the observed series.
3. Fit an **ARIMA** model (via `auto.arima`) with **no drift**.
4. **Simulate** many synthetic series from the fitted ARIMA model.
5. For each simulated series, compute Sen’s slope (per decade).
6. Use the distribution of simulated slopes to define **Q1-Q7** quantile bins.
7. Place the observed slope into one of the bins (**qcat**).

The resulting `qcat` is a standardized way to summarize whether the observed trend is unusually strong relative to what would be expected under the fitted stationary ARIMA process.

## Load the generated CSV outputs

```{r read-arima-csv}
all_path <- file.path(out_dir, "FIRE_ALL_questions_ARIMA_outputs.csv")
summary_path <- file.path(out_dir, "FIRE_metric_summary_by_Qcat.csv")

stopifnot(file.exists(all_path))

df_all <- readr::read_csv(all_path, show_col_types = FALSE)

# ------------------------------
# ARIMA / Delta diagnostics: more visual summaries
# ------------------------------
# This block tries to create a few high-signal figures from the ARIMA outputs.
# It is written defensively: if a needed column is missing, the figure is skipped.

# Ensure question is a clean ordered factor if it exists (Q1..Q5 expected here)
if ("question" %in% names(df_all)) {
  q_levels <- sort(unique(df_all$question))
  if (all(grepl("^Q[0-9]+$", q_levels))) {
    q_levels <- paste0("Q", sort(as.integer(sub("Q", "", q_levels))))
  }
  df_all$question <- factor(df_all$question, levels = q_levels)
}

# Convert qcat to an ordered factor if present
if ("qcat" %in% names(df_all)) {
  qcat_levels <- paste0("Q", 1:7)
  df_all$qcat <- factor(df_all$qcat, levels = qcat_levels, ordered = TRUE)
  df_all$qcat_num <- as.integer(df_all$qcat)
}

# A compact unit id for plotting (town + buffer)
if (all(c("town","buffer") %in% names(df_all))) {
  df_all$unit <- paste(df_all$town, df_all$buffer, sep = " | ")
}

df_sum <- readr::read_csv(summary_path, show_col_types = FALSE)

cat("(Initial rows table suppressed.)\n")
```

## ARIMA results: visual summaries

```{r arima-qcat-heatmap}
# Heatmap of Qcat by unit (town|buffer) and question
if (all(c("qcat","unit","question") %in% names(df_all))) {
  p_heat <- ggplot(df_all, aes(x = question, y = unit, fill = qcat_num)) +
    geom_tile(color = "white", linewidth = 0.2) +
    scale_fill_continuous(breaks = 1:7, limits = c(1,7)) +
    labs(
      title = "ARIMA Delta diagnostics: Q1-Q7 category by question and location",
      x = "Question",
      y = "Town | Buffer",
      fill = "Qcat (1-7)"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.y = element_text(size = 7),
      panel.grid = element_blank()
    )
  p_heat
} else {
  cat("(Heatmap skipped: need columns question, town, buffer, and qcat in FIRE_ALL_questions_ARIMA_outputs.csv.)\n")
}
```

```{r arima-qcat-counts}
# Distribution of Qcat by question
if (all(c("qcat","question") %in% names(df_all))) {
  p_bar <- df_all %>%
    count(question, qcat) %>%
    ggplot(aes(x = question, y = n, fill = qcat)) +
    geom_col(position = "stack") +
    labs(
      title = "Counts of ARIMA Qcat categories by question",
      x = "Question",
      y = "Count",
      fill = "Qcat"
    ) +
    theme_minimal(base_size = 11) +
    scale_fill_discrete(drop = FALSE)
  p_bar
} else {
  cat("(Bar chart skipped: need columns question and qcat.)\n")
}
```

```{r arima-question-key}
question_key <- tibble::tibble(
  question = c("Q1","Q2","Q3","Q4","Q5"),
  title = c(
    "Overall frequency of high-fire conditions",
    "Summer frequency of high-fire conditions (JJA)",
    "Fire severity / intensity proxy (large or impactful fire-weather potential)",
    "Fire season start timing",
    "Fire season duration / persistence"
  ),
  survey_question = c(
    "Overall, has fire become less common or more common?",
    "During the summer months, has fire occurred more often or less often?",
    "Have the fires been smaller or larger / less or more severe?",
    "Has there been a change in when the fire season starts?",
    "Do the fire conditions last shorter or longer than you remember?"
  )
)

knitr::kable(
  question_key,
  caption = "Survey question key for the question buckets used in the ARIMA summary plots.",
  align = c("l","l","l")
)
```


```{r arima-delta-by-qcat}
# If an observed delta/trend magnitude column exists, show it by Qcat
# We'll look for a few common names and use the first one found.
delta_col <- intersect(c("delta","observed_delta","sen_slope","sen_slope_C_per_decade","slope"), names(df_all))
if (length(delta_col) > 0 && all(c("qcat","question") %in% names(df_all))) {
  dc <- delta_col[1]
  p_box <- ggplot(df_all, aes(x = qcat, y = .data[[dc]])) +
    geom_boxplot(outlier.alpha = 0.4) +
    facet_wrap(~ question, scales = "free_y") +
    labs(
      title = paste0("Observed trend magnitude vs ARIMA Qcat (using ", dc, ")"),
      x = "Qcat",
      y = dc
    ) +
    theme_minimal(base_size = 11)
  p_box
} else {
  cat("(Boxplot skipped: need qcat + question and a numeric trend column such as observed_delta or sen_slope_C_per_decade.)\n")
}
```


## Summary counts by question, metric, and qcat

```{r arima-summary-table, include=FALSE}
# Full summary table suppressed to avoid printing large tables in the PDF.
# The CSV is saved to outputs/fire_questions/FIRE_metric_summary_by_Qcat.csv
```

*(ARIMA summary table suppressed; see saved CSV outputs.)*

## Quick sanity checks

```{r arima-sanity, include=FALSE}
# Sanity-check tables suppressed to keep the report clean.
```


# Notes / interpretation guidance

- **Frequency metrics** (Q1/Q2) increasing typically indicate more frequent high fire-weather conditions.
- **Timing metrics** (Q4) decreasing (more negative Sen slope) indicates *earlier* onset over time because wyday gets smaller.
- **Magnitude proxies** (Q3) are not burned area, but can align with years of high potential fire behavior.
- **Qcat values** are **relative to simulated ARIMA variability**, not a direct p-value test of a trend model.


# Estimating PDI (Perception Δ Instrumentation)

This section links the **ARIMA-derived instrumentation signal** to the **survey responses** *without using respondent tenure*.
Conceptually:

- **Perception**: survey response for each question (Likert 1-5; higher = "more / earlier / longer / more severe" depending on wording)
- **Instrumentation**: ARIMA-derived trend bucket (**Q1-Q7**) or continuous **Sen slope per decade** for the corresponding fire metrics
- **PDI**: mismatch between perception and instrumentation after putting them on comparable scales.

## Load ARIMA outputs + survey responses

```{r pdi-load, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)

# Prefer project-relative paths (as in your repo), but fall back to local files if needed.
arima_file <- "outputs/fire_questions/FIRE_ALL_questions_ARIMA_outputs.csv"
survey_file <- "data/survey/FW2L_DC.csv"

if (!file.exists(arima_file)) arima_file <- "FIRE_ALL_questions_ARIMA_outputs.csv"
if (!file.exists(survey_file)) survey_file <- "FW2L_DC.csv"

arima <- readr::read_csv(arima_file, show_col_types = FALSE)
survey <- readr::read_csv(survey_file, show_col_types = FALSE)

# Harmonize town naming (ARIMA: ElkRiver; survey: Elk River)
survey <- survey %>%
  mutate(town = case_when(
    Nearest_Town == "Elk River" ~ "ElkRiver",
    TRUE ~ Nearest_Town
  ))

arima <- arima %>%
  mutate(town = as.character(town))
```



```{r exclude-troy, include=FALSE}
library(dplyr)
library(stringr)

drop_town <- function(x) str_to_lower(str_trim(as.character(x)))

# Exclude Troy everywhere (survey + ARIMA outputs)
survey <- survey %>%
  filter(drop_town(town) != "troy")

arima <- arima %>%
  filter(drop_town(town) != "troy")

# If you have annual tables already built, exclude there too
if (exists("annualfire_long")) {
  annualfire_long <- annualfire_long %>%
    filter(drop_town(town) != "troy")
}

if (exists("annualfire_long2")) {
  annualfire_long2 <- annualfire_long2 %>%
    filter(drop_town(town) != "troy")
}
```






```{r pdi-tenure-setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)

# ---- Map survey items to questions (you provided this) ----
survey_q_map <- tribble(
  ~question, ~survey_col, ~direction,
  "Q1",      "ID22",      "higher_more",
  "Q2",      "ID23",      "higher_more",
  "Q3",      "ID24",      "higher_more",
  "Q4",      "ID25",      "higher_more",
  "Q5",      "ID26",      "higher_more"
)

# ---- Map questions to the instrumentation metrics you use ----
# IMPORTANT: metric names must match annual_long$metric exactly.
q_metric_map <- tribble(
  ~question, ~metric,
  "Q1", "firehighdays_erc90",
  "Q1", "firehighdays_bi90",

  "Q2", "firesummerhighdays_erc90",
  "Q2", "firesummerhighdays_bi90",

  "Q3", "firepeak_erc",
  "Q3", "firepeak_bi",
  "Q3", "firep95_erc",

  "Q4", "firestart_wyday_erc",
  "Q4", "firestart_wyday_bi",

  "Q5", "fireseasonlength_erc",
  "Q5", "firemaxrun_erc",
  "Q5", "firemaxrun_bi"
)

to_num <- function(x) suppressWarnings(as.numeric(x))

compute_delta_one_series <- function(df_one, end_year, tenure_years,
                                     min_years = 8,
                                     delta_method = c("slope","diff"),
                                     k_years = 5) {
  delta_method <- match.arg(delta_method)
  if (is.na(tenure_years)) return(NA_real_)

  df_one <- df_one %>% filter(!is.na(value))
  if (nrow(df_one) == 0) return(NA_real_)

  max_year <- min(end_year, max(df_one$year, na.rm = TRUE))
  start_year <- max_year - floor(tenure_years) + 1

  df_w <- df_one %>%
    filter(year >= start_year, year <= max_year) %>%
    arrange(year)

  if (nrow(df_w) < min_years) return(NA_real_)

  if (delta_method == "slope") {
    fit <- try(lm(value ~ year, data = df_w), silent = TRUE)
    if (inherits(fit, "try-error")) return(NA_real_)
    slope_per_year <- unname(coef(fit)[["year"]])
    return(10 * slope_per_year)  # per decade
  }

  # delta_method == "diff"
  k <- min(k_years, floor(nrow(df_w) / 2))
  if (k < 2) return(NA_real_)
  first_mean <- mean(df_w$value[1:k], na.rm = TRUE)
  last_mean  <- mean(df_w$value[(nrow(df_w)-k+1):nrow(df_w)], na.rm = TRUE)
  last_mean - first_mean
}

compute_pdi_tenure <- function(survey,
                               annual_long,
                               survey_q_map,
                               q_metric_map,
                               town_col = "town",
                               id_col   = "record_id",
                               tenure_col = "tenure_years",
                               end_year = 2025,
                               delta_method = c("slope","diff"),
                               k_years = 5,
                               min_years = 8) {
  delta_method <- match.arg(delta_method)

  perception_long <- survey %>%
    mutate(
      .town = .data[[town_col]],
      .id   = .data[[id_col]],
      tenure_years = to_num(.data[[tenure_col]])
    ) %>%
    select(.id, .town, tenure_years, all_of(survey_q_map$survey_col)) %>%
    pivot_longer(
      cols = all_of(survey_q_map$survey_col),
      names_to = "survey_item",
      values_to = "resp"
    ) %>%
    left_join(survey_q_map, by = c("survey_item" = "survey_col")) %>%
    mutate(
      resp = to_num(resp),
      resp = if_else(direction == "reverse" & !is.na(resp), 8 - resp, resp)  # 1–7 Likert
    ) %>%
    select(.id, .town, tenure_years, question, resp)

  person_q_metric <- perception_long %>%
    left_join(q_metric_map, by = "question") %>%
    filter(!is.na(metric))

  stopifnot(all(c("town","year","metric","value") %in% names(annual_long)))

  deltas <- person_q_metric %>%
    left_join(annual_long, by = c(".town" = "town", "metric" = "metric")) %>%
    group_by(.id, .town, tenure_years, question, resp, metric) %>%
    summarise(
      delta = compute_delta_one_series(
        df_one = tibble(year = year, value = value),
        end_year = end_year,
        tenure_years = unique(tenure_years),
        min_years = min_years,
        delta_method = delta_method,
        k_years = k_years
      ),
      .groups = "drop"
    )

  instr_person_q <- deltas %>%
    group_by(.id, .town, tenure_years, question, resp) %>%
    summarise(
      instr_delta = mean(delta, na.rm = TRUE),
      n_metrics_used = sum(!is.na(delta)),
      .groups = "drop"
    )

  instr_person_q %>%
    group_by(question) %>%
    mutate(
      z_resp  = as.numeric(scale(resp)),
      z_delta = as.numeric(scale(instr_delta)),
      PDI = z_resp - z_delta
    ) %>%
    ungroup()
}
```


## Map survey items to questions (edit this once)

You need to tell the pipeline which survey columns correspond to each fire-perception question **Q1-Q5**.
Set this mapping once, then everything below will run.

```{r pdi-map, include=FALSE}
library(tibble)
library(dplyr)

# Mapping of the 5 fire-perception survey items (provided by you)
# NOTE: ID22–ID26 are 1–7 Likert items in FW2L_DC.csv
survey_q_map <- tribble(
  ~question, ~survey_col, ~direction,
  "Q1",      "ID22",      "higher_more",
  "Q2",      "ID23",      "higher_more",
  "Q3",      "ID24",      "higher_more",
  "Q4",      "ID25",      "higher_more",
  "Q5",      "ID26",      "higher_more"
)

# If any item is reverse-coded (higher means "less / later / shorter"),
# change direction for that row to "reverse" and it will be flipped automatically (1<->7).
flip_likert_1_7 <- function(x) {
  # works for numeric or integer; preserves NA
  ifelse(is.na(x), NA, 8 - as.numeric(x))
}
```

## Compute PDI (town-level, without tenure)

We compute:

1. **Perception score** per town × question: mean Likert response (after optional reverse-coding)  
2. **Instrumentation score** per town × question: mean ARIMA bucket (Q1-Q7 -> 1..7) across that question's metrics  
3. **PDI** per town × question: z(perception) − z(instrumentation), standardized within question

```{r pdi-compute, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)

# ---- Perception: town × question mean response ----
perception <- survey %>%
  # keep just what we need
  select(town, all_of(survey_q_map$survey_col)) %>%
  pivot_longer(
    cols = all_of(survey_q_map$survey_col),
    names_to = "survey_item",
    values_to = "resp"
  ) %>%
  # attach question + direction
  left_join(
    survey_q_map %>% mutate(survey_col = as.character(survey_col)),
    by = c("survey_item" = "survey_col")
  ) %>%
  mutate(
    resp = suppressWarnings(as.numeric(resp)),
    # reverse-code for 1–7 Likert if needed
    resp = if_else(direction == "reverse" & !is.na(resp), 8 - resp, resp)
  ) %>%
  group_by(town, question) %>%
  summarise(
    perception_mean = mean(resp, na.rm = TRUE),
    perception_n = sum(!is.na(resp)),
    .groups = "drop"
  )

# ---- Instrumentation: town × question mean ARIMA bucket across metrics ----
instrumentation <- arima %>%
  mutate(
    qcat_num = as.integer(gsub("^Q","", qcat)),
    qcat_num = if_else(is.na(qcat_num), NA_integer_, qcat_num)
  ) %>%
  group_by(town, question, buffer) %>%
  summarise(
    instr_qcat_mean = mean(qcat_num, na.rm = TRUE),
    instr_sen_slope_mean = mean(sen_slope_per_decade, na.rm = TRUE),
    n_metrics = n(),
    .groups = "drop"
  )

# Choose which buffer to use for PDI (default: none)
buffer_choice <- "none"

instrumentation_use <- instrumentation %>%
  filter(buffer == buffer_choice) %>%
  select(town, question, instr_qcat_mean, instr_sen_slope_mean, n_metrics)

# ---- Join and compute PDI ----
pdi <- perception %>%
  left_join(instrumentation_use, by = c("town","question")) %>%
  group_by(question) %>%
  mutate(
    z_perception = as.numeric(scale(perception_mean)),
    z_instr = as.numeric(scale(instr_qcat_mean)),
    PDI = z_perception - z_instr
  ) %>%
  ungroup()

pdi
```

### PDI summary table

```{r pdi-table}
pdi %>%
  arrange(question, town) %>%
  select(town, question, perception_mean, perception_n, instr_qcat_mean, n_metrics, PDI) %>%
  knitr::kable(digits = 3)
```

### PDI plot (by town)

```{r pdi-plot, fig.height=4.2}
pdi %>%
  ggplot(aes(x = town, y = PDI)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point(size = 2) +
  facet_wrap(~ question, ncol = 2) +
  coord_cartesian(ylim = range(pdi$PDI, na.rm = TRUE)) +
  labs(
    title = "PDI (z-perception − z-instrumentation) by town and question",
    x = NULL,
    y = "PDI"
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```


```{r pdi-tenure-annual, include=FALSE}
# Use the annual town×year×metric table created by extract_fire_metrics2-2.R
# annualfire_long columns: town, buffer, wateryear, metric, value

buffer_choice_person <- "none"  # or "10mibuffer"

annual_long <- annualfire_long %>%
  filter(buffer == buffer_choice_person) %>%
  transmute(
    town = as.character(town),
    year = as.integer(wateryear),
    metric = as.character(metric),
    value = as.numeric(value)
  )
```



```{r pdi-tenure-compute, message=FALSE, warning=FALSE}
# >>> CHANGE THESE THREE to match your survey columns <<<
TOWN_COL   <- "Nearest_Town"
ID_COL     <- "ObjectID"
TENURE_COL <- "ID2"

# Compute individualized instrumentation Delta (per person×question)
# and PDI = z(response) - z(delta), standardized within question
pdi_person <- compute_pdi_tenure(
  survey = survey,
  annual_long = annual_long,
  survey_q_map = survey_q_map,
  q_metric_map = q_metric_map,
  town_col = TOWN_COL,
  id_col = ID_COL,
  tenure_col = TENURE_COL,
  end_year = 2025,
  delta_method = "slope",  # "slope" (per decade) or "diff"
  k_years = 5,
  min_years = 8
)

```



```{r pdi-tenure-plot-alltowns, fig.height=7, fig.width=7}
ggplot(pdi_person, aes(x = resp, y = instr_delta)) +
  geom_point(alpha = 0.35) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ question, scales = "free_y") +
  labs(
    x = "Survey response (Likert 1–7)",
    y = "Instrumentation Delta (tenure-based; slope per decade)",
    title = "Perception vs Individualized Instrumentation Delta (All towns)"
  )
```


```{r pdi-tenure-plot-bytown, fig.height=10, fig.width=8}
ggplot(pdi_person, aes(x = resp, y = instr_delta)) +
  geom_point(alpha = 0.35) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(question ~ .town, scales = "free_y") +
  labs(
    x = "Survey response (Likert 1–7)",
    y = "Instrumentation Delta (tenure-based; slope per decade)",
    title = "Perception vs Individualized Instrumentation Delta (By town)"
  )
```
  
# Keep this chunk quiet in the PDF; it's here for debugging if needed
# dplyr::glimpse(pdi_person)

## Town demographics (ACS block-group population-weighted buffers)

This section derives town-level demographic covariates by intersecting a town-centered buffer with Census **block groups** and computing **population-weighted** averages of ACS measures within the buffer.  
**You must provide town coordinates (lon/lat)** below. If two towns fall within the same tract/block group, this approach can still differ across towns because it aggregates across the *buffer footprint* rather than assigning a single geography.

### 8.1 Setup + town coordinates (edit these)

```{r acs-setup, include=FALSE}
library(sf)
library(dplyr)
library(tidyr)
library(tibble)
library(tidycensus)
library(tigris)
options(tigris_use_cache = TRUE)

# ---- Census API key ----
# Recommended: set once in your environment (e.g., .Renviron) rather than in the Rmd:
# census_api_key("YOUR_KEY_HERE", install = TRUE)

# ---- Town coordinates (loaded from CSV) ----
# Expected file columns (case-insensitive):
#   - town (or city)
#   - lon (or longitude)
#   - lat (or latitude)
#
# Preferred project location: data/coordinates/city_coordinates.csv
# (During ChatGPT runs, a copy may exist at /mnt/data/city_coordinates.csv)
coord_path <- dplyr::case_when(
  file.exists('data/coordinates/city_coordinates.csv') ~ 'data/coordinates/city_coordinates.csv',
  file.exists('/mnt/data/city_coordinates.csv') ~ '/mnt/data/city_coordinates.csv',
  TRUE ~ NA_character_
)

if (is.na(coord_path)) {
  stop('Could not find city_coordinates.csv at data/coordinates/city_coordinates.csv (preferred) or /mnt/data/city_coordinates.csv.')
}

town_coords_raw <- readr::read_csv(coord_path, show_col_types = FALSE)

# Normalize column names
nm <- tolower(names(town_coords_raw))
names(town_coords_raw) <- nm

# Allow either 'town' or 'city' for the name field
if (!('town' %in% nm) && ('city' %in% nm)) {
  town_coords_raw <- town_coords_raw %>% dplyr::rename(town = city)
}

# Allow lon/lat aliases
if (!('lon' %in% names(town_coords_raw)) && ('longitude' %in% names(town_coords_raw))) {
  town_coords_raw <- town_coords_raw %>% dplyr::rename(lon = longitude)
}
if (!('lat' %in% names(town_coords_raw)) && ('latitude' %in% names(town_coords_raw))) {
  town_coords_raw <- town_coords_raw %>% dplyr::rename(lat = latitude)
}

# Filter to towns used in this report (edit if you change towns)
towns_needed <- c('Bovill','Deary','Elk River','Juliaetta','Kendrick')

town_coords <- town_coords_raw %>%
  dplyr::mutate(
    town = as.character(town),
    lon = suppressWarnings(as.numeric(lon)),
    lat = suppressWarnings(as.numeric(lat))
  ) %>%
  dplyr::filter(town %in% towns_needed) %>%
  dplyr::select(town, lon, lat) %>%
  dplyr::distinct()

# Check that we found all towns and have valid coords
missing_towns <- setdiff(towns_needed, town_coords$town)
if (length(missing_towns) > 0) {
  stop(paste0('Missing coordinates for: ', paste(missing_towns, collapse = ', '),
              '. Check city_coordinates.csv town/city names match exactly.'))
}
if (any(is.na(town_coords$lon)) || any(is.na(town_coords$lat))) {
  stop('Found NA lon/lat in city_coordinates.csv after filtering. Fix the file and re-knit.')
}

stopifnot(all(!is.na(town_coords$town)))
# If you forget to fill coords, error early so you notice:
if (any(is.na(town_coords$lon)) || any(is.na(town_coords$lat))) {
  stop("Fill in lon/lat for all towns in town_coords before knitting.")
}

# Keep town names consistent with the rest of the report
town_coords <- town_coords %>% mutate(town = as.character(town))

town_sf <- st_as_sf(town_coords, coords = c("lon","lat"), crs = 4326)

# Use a projected CRS for accurate buffering/areas in meters.
# EPSG:3857 is okay for small-area buffers; for Idaho, EPSG:26911/26912 can also work.
town_sf_m <- st_transform(town_sf, 3857)
```

### 8.2 Choose variables + pull block groups

```{r acs-vars, include=FALSE}
# ---- Choose ACS year (5-year) ----
acs_year <- 2022

# ---- Choose state for Census geographies (Idaho = "ID") ----
acs_state <- "ID"

# ---- Buffer for demographics (miles) ----
# For "social context" 1–3 miles is often better than 10 miles in rural areas.
demo_buffer_miles <- 3
demo_buffer_m <- demo_buffer_miles * 1609.34

# ---- ACS variables (keep compact for manuscript) ----
# Total population
acs_vars <- c(
  pop_total  = "B01003_001",
  med_income = "B19013_001",
  # poverty: below poverty / poverty universe
  pov_total  = "B17001_001",
  pov_below  = "B17001_002",
  # education: bachelor's+ counts (sum of these)
  bach_plus  = "B15003_022",
  mast_plus  = "B15003_023",
  prof_plus  = "B15003_024",
  doc_plus   = "B15003_025"
)

# Age 65+ requires summing multiple B01001 cells (male 65+ and female 65+)
age65_vars <- c(
  paste0("B01001_", sprintf("%03d", 20:25)),  # male 65+ bins
  paste0("B01001_", sprintf("%03d", 44:49))   # female 65+ bins
)
```

### 8.3 Compute population-weighted buffer covariates (block groups)

```{r acs-buffer-aggregate, message=FALSE, warning=FALSE}
# ---- Buffers ----
town_buf <- st_buffer(town_sf_m, dist = demo_buffer_m) %>%
  dplyr::select(town)

# ---- Block groups geometry ----
bg <- tigris::block_groups(state = acs_state, year = acs_year, cb = TRUE, class = "sf") %>%
  st_transform(st_crs(town_buf))

# ---- ACS data at block-group level ----
acs_bg_long <- get_acs(
  geography = "block group",
  variables = c(acs_vars, setNames(age65_vars, age65_vars)),
  state = acs_state,
  year = acs_year,
  survey = "acs5",
  geometry = FALSE,
  cache_table = TRUE
)

acs_bg <- acs_bg_long %>%
  dplyr::select(GEOID, variable, estimate) %>%
  tidyr::pivot_wider(names_from = variable, values_from = estimate)

# ---- Derived measures ----
acs_bg <- acs_bg %>%
  mutate(
    # education counts and rates
    bachplus_n = bach_plus + mast_plus + prof_plus + doc_plus,
    bachplus_rate = if_else(pop_total > 0, bachplus_n / pop_total, NA_real_),
    # poverty rate
    pov_rate = if_else(pov_total > 0, pov_below / pov_total, NA_real_),
    # 65+ rate
    age65_n = rowSums(across(all_of(age65_vars)), na.rm = TRUE),
    age65_rate = if_else(pop_total > 0, age65_n / pop_total, NA_real_)
  ) %>%
  select(GEOID, pop_total, med_income, pov_rate, bachplus_rate, age65_rate)

bg_acs <- bg %>% left_join(acs_bg, by = "GEOID")

# ---- Intersect buffers with BGs ----
# We weight BG contributions by estimated population inside the buffer:
#   pop_wt = pop_total * (intersection_area / bg_area)
int <- st_intersection(
  town_buf,
  bg_acs %>% select(GEOID, pop_total, med_income, pov_rate, bachplus_rate, age65_rate)
)

# ---- Compute weights ----
bg_area_tbl <- bg_acs %>%
  mutate(bg_area = as.numeric(st_area(geometry))) %>%
  st_drop_geometry() %>%
  select(GEOID, bg_area)

int <- int %>%
  mutate(int_area = as.numeric(st_area(geometry))) %>%
  left_join(bg_area_tbl, by = "GEOID") %>%
  mutate(
    area_wt = if_else(bg_area > 0, int_area / bg_area, NA_real_),
    pop_wt  = pop_total * area_wt
  )

# ---- Town-level covariates (population-weighted) ----
town_covars <- int %>%
  st_drop_geometry() %>%
  group_by(town) %>%
  summarise(
    pop_in_buffer = sum(pop_wt, na.rm = TRUE),
    med_income_wt = weighted.mean(med_income, w = pop_wt, na.rm = TRUE),
    pov_rate_wt   = weighted.mean(pov_rate,   w = pop_wt, na.rm = TRUE),
    age65_rate_wt = weighted.mean(age65_rate, w = pop_wt, na.rm = TRUE),
    bachplus_rate_wt = weighted.mean(bachplus_rate, w = pop_wt, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(town)

town_covars
```

### 8.4 Diagnostics: shared geographies and buffer overlap

```{r acs-diagnostics, message=FALSE, warning=FALSE}
# Do towns fall within the same containing block group?
town_bg <- st_join(town_sf_m %>% select(town), bg_acs %>% select(GEOID), join = st_within) %>%
  st_drop_geometry()

town_bg %>% count(GEOID) %>% arrange(desc(n))

# How much do buffers overlap (percent of the smaller buffer's area)?
buf_pairs <- tidyr::expand_grid(t1 = town_buf$town, t2 = town_buf$town) %>%
  filter(t1 < t2)

overlap_tbl <- purrr::pmap_dfr(buf_pairs, function(t1, t2) {
  g1 <- town_buf %>% filter(town == t1)
  g2 <- town_buf %>% filter(town == t2)
  a1 <- as.numeric(st_area(g1))
  a2 <- as.numeric(st_area(g2))
  ai <- as.numeric(st_area(st_intersection(g1, g2)))
  tibble(t1 = t1, t2 = t2, overlap_pct_of_smaller = 100 * ai / min(a1, a2))
})

overlap_tbl %>% arrange(desc(overlap_pct_of_smaller))
```

### 8.5 Join demographics to PDI + manuscript-ready plots

```{r pdi-join-demographics, message=FALSE, warning=FALSE}
# Join to your town-level PDI table (ARIMA-based PDI)
# Assumes pdi has columns: town, question, PDI
pdi_demo <- pdi %>%
  left_join(town_covars, by = "town")

pdi_demo
```

```{r pdi-heatmap-with-demo, fig.height=5.5, fig.width=7.5}
# Heatmap of PDI (town × question)
ggplot(pdi_demo, aes(x = question, y = town, fill = PDI)) +
  geom_tile(color = "white", linewidth = 0.2) +
  geom_text(aes(label = ifelse(is.na(PDI), "", sprintf("%.2f", PDI))), size = 3) +
  labs(
    title = "Town × Question PDI (ARIMA-based)",
    subtitle = paste0("Demographics buffer: ", demo_buffer_miles, " miles (ACS ", acs_year, " block groups)"),
    x = "Question", y = "Town", fill = "PDI"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

```{r pdi-vs-income, fig.height=5.5, fig.width=7.5}
# Town mean PDI (across questions) vs demographics (example)
# Requires ggrepel
if (!requireNamespace("ggrepel", quietly = TRUE)) {
  stop("Please install ggrepel (install.packages('ggrepel')) to run this plot.")
}

town_mean_pdi <- pdi_demo %>%
  group_by(town) %>%
  summarise(mean_PDI = mean(PDI, na.rm = TRUE), .groups = "drop") %>%
  left_join(town_covars, by = "town")

ggplot(town_mean_pdi, aes(x = med_income_wt, y = mean_PDI, label = town)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel() +
  labs(
    title = "Town mean PDI vs median household income (ACS-weighted buffer)",
    x = "Median household income (population-weighted within buffer)",
    y = "Mean PDI (across Q1–Q5)"
  ) +
  theme_minimal()
```


## Notes

- This is a **tenure-free PDI**: instrumentation reflects the *place’s recent trajectory* (ARIMA/Qcat), not each respondent’s personal exposure window.
- If you want an **individual-level** PDI without tenure, we can still do it: each respondent inherits their town’s instrumentation score, then PDI is computed per respondent and summarized by town/demographics.
