---
title: "Idaho Climate Metrics Report (By Question)"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
geometry:
  - margin=1in
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{graphicx}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6.5,
  fig.height = 4.2,
  dpi = 300,
  fig.align = 'center',
  out.width = '100%'
)
```

# Overview

This report produces a **question-framed** summary of **heat, drought, and fire** metrics for the five Idaho towns:

- Deary
- Bovill
- Kendrick
- Juliaetta
- Elk River

The workflow mirrors the original fire report structure as closely as possible, while extending the analysis to the **climate-comparable survey questions** for heat, drought, and fire.

The pipeline:

1. Reads daily **GridMET** time series by town and buffer.
2. Builds **daily climatologies** (1996-2025) and seasonal or threshold-based daily flags.
3. Aggregates daily values into **annual water-year metrics**.
4. Maps annual metrics into **survey-question buckets**.
5. Computes **Delta / ARIMA** diagnostics for each metric.
6. Computes **tenure-based PDI** summaries from the linked survey responses.

This RMarkdown is designed so that **knitting the PDF runs the full pipeline once** by sourcing the working Idaho script, then renders plots and summary tables from the generated objects and saved outputs.

# Data sources and conventions

## Data source

- **GridMET** daily climate time series for each town and its 10-mile buffer.
- Expected input files follow the existing project naming convention:
  - `Town_gridmet.csv`
  - `Town_10mibuffer_gridmet.csv`

## Calendar and water-year conventions

- **Water year**: Oct 1 - Sep 30, labeled by the ending year.
- **wyday**: water-year day where Oct 1 = 1.
- **season4**: DJF / MAM / JJA / SON.
- **baseline climatology period**: 1996-2025.

```{r source-pipeline, include=FALSE}
library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(tidyr)
library(lubridate)
library(zoo)
library(ggplot2)
library(trend)
library(forecast)
library(knitr)

project_root <- getwd()

script_candidates <- file.path(project_root, "src", "extract_idaho_heat_drought_fire_metrics_v2.R")

script_path <- script_candidates[file.exists(script_candidates)][1]
stopifnot(!is.na(script_path), file.exists(script_path))

plot_dir <- file.path(project_root, "plots")
out_dir  <- file.path(project_root, "outputs")

source(script_path)
```

# Metric construction and question mapping

## Heat questions

- **H1**: Overall heat magnitude
- **H2**: Summer heat magnitude
- **H3**: Heat frequency
- **H4**: Spring heat onset timing
- **H5**: Fall cooling timing
- **H6**: Heat duration / persistence

## Drought questions

- **D1**: Overall wetter/drier magnitude
- **D2**: Winter wetter/drier magnitude
- **D3**: Winter dryness frequency
- **D4**: Precipitation timing
- **D5**: Dry-duration persistence

## Fire questions

- **F1**: Overall frequency of high fire-weather days
- **F2**: Summer frequency of high fire-weather days
- **F3**: Fire-weather magnitude / intensity proxy
- **F4**: Fire-season start timing
- **F5**: Fire-season duration / persistence

```{r read-generated-csvs, include=FALSE}
question_outputs_path <- file.path(out_dir, "idaho_question_outputs.csv")
pdi_summary_path      <- file.path(out_dir, "idaho_pdi_summary.csv")
annual_long_path      <- file.path(out_dir, "idaho_annual_metrics_long.csv")

stopifnot(file.exists(question_outputs_path))
stopifnot(file.exists(pdi_summary_path))
stopifnot(file.exists(annual_long_path))

question_outputs <- readr::read_csv(question_outputs_path, show_col_types = FALSE)
pdi_summary      <- readr::read_csv(pdi_summary_path, show_col_types = FALSE)
annual_long_csv  <- readr::read_csv(annual_long_path, show_col_types = FALSE)

question_meta <- tibble::tribble(
  ~question, ~stressor, ~title, ~survey_question, ~interpretation,
  "H1", "Heat", "Overall heat magnitude", "Overall, has it gotten hotter or cooler?", "Higher values indicate stronger overall warming relative to baseline conditions.",
  "H2", "Heat", "Summer heat magnitude", "During the summer months, has it gotten hotter or cooler?", "Higher values indicate hotter summer conditions, especially in JJA.",
  "H3", "Heat", "Heat frequency", "If summer has changed, how frequently is it hotter?", "Higher values indicate more frequent hot-day conditions.",
  "H4", "Heat", "Spring heat onset timing", "When does it start getting hotter in the spring?", "Smaller day-of-year values indicate earlier spring onset.",
  "H5", "Heat", "Fall cooling timing", "When does it start getting cooler in the fall?", "Larger end-day values indicate later persistence of warm-season conditions.",
  "H6", "Heat", "Heat duration / persistence", "Do heat conditions last shorter or longer than you remember?", "Higher values indicate longer hot periods or growing-season persistence.",
  "D1", "Drought", "Overall wetter/drier magnitude", "Overall, has it gotten wetter or drier?", "Lower precipitation / larger moisture deficits indicate stronger drought stress.",
  "D2", "Drought", "Winter wetter/drier magnitude", "During the winter months, has it become wetter or drier?", "Focuses on cool-season precipitation conditions.",
  "D3", "Drought", "Winter dryness frequency", "If winter has changed, how frequently is it drier?", "Higher values indicate more frequent winter dry conditions.",
  "D4", "Drought", "Precipitation timing", "Has there been a change in when rainfall/snowfall starts during the year?", "Timing metrics summarize onset and center-of-mass shifts in precipitation.",
  "D5", "Drought", "Dry-duration persistence", "Do drier conditions last shorter or longer than you remember?", "Higher values indicate longer dry spells.",
  "F1", "Fire", "Overall frequency of high fire-weather days", "Overall, has fire become less common or more common?", "Higher values indicate more days with unusually elevated fire-weather conditions.",
  "F2", "Fire", "Summer frequency of high fire-weather days", "During the summer months, has fire occurred more often or less often?", "Higher values indicate more frequent summer fire-weather exceedances.",
  "F3", "Fire", "Fire-weather magnitude / intensity proxy", "Have the areas affected by wildfires gotten smaller or larger?", "Higher values indicate more extreme or intense fire-weather conditions.",
  "F4", "Fire", "Fire-season start timing", "Has there been a change in when the fire season starts during the year?", "Smaller water-year-day values indicate earlier fire-season onset.",
  "F5", "Fire", "Fire-season duration / persistence", "Do the fire conditions last shorter or longer than you remember?", "Higher values indicate longer or more persistent fire-weather periods."
)
```

# Results: Question-framed plots

```{r helper-include-plot, include=FALSE}
include_plot <- function(fname) {
  f <- file.path(plot_dir, fname)
  if (!file.exists(f)) {
    cat("(Plot not found:", fname, ")\n")
    return(invisible(NULL))
  }
  knitr::include_graphics(f)
}

show_question_section <- function(qcode) {
  meta <- question_meta %>% filter(question == qcode)
  qdat <- question_outputs %>% filter(question == qcode) %>% arrange(metric, town, buffer)
  metrics <- unique(qdat$metric)

  cat("**Survey question:** ", meta$survey_question, "\n\n", sep = "")
  cat("**Interpretation:** ", meta$interpretation, "\n\n", sep = "")
  cat("**Metrics used:**\n")
  for (m in metrics) cat("- `", m, "`\n", sep = "")
  cat("\n")

  for (m in metrics) {
    include_plot(paste0(qcode, "_timeseries_", m, ".png"))
  }
  for (m in metrics) {
    include_plot(paste0(qcode, "_distribution_", m, ".png"))
  }

  tbl <- qdat %>%
    select(town, buffer, metric, sen_slope_per_decade, qcat, delta_late_minus_early) %>%
    mutate(across(c(sen_slope_per_decade, delta_late_minus_early), ~ round(.x, 3)))

  if (nrow(tbl) > 0) {
    print(knitr::kable(tbl, caption = paste(qcode, meta$title), align = c("l", "l", "l", "r", "l", "r")))
  }
}
```

# Heat

## H1: Overall heat magnitude

```{r H1}
show_question_section("H1")
```

## H2: Summer heat magnitude

```{r H2}
show_question_section("H2")
```

## H3: Heat frequency

```{r H3}
show_question_section("H3")
```

## H4: Spring heat onset timing

```{r H4}
show_question_section("H4")
```

## H5: Fall cooling timing

```{r H5}
show_question_section("H5")
```

## H6: Heat duration / persistence

```{r H6}
show_question_section("H6")
```

# Drought

## D1: Overall wetter/drier magnitude

```{r D1}
show_question_section("D1")
```

## D2: Winter wetter/drier magnitude

```{r D2}
show_question_section("D2")
```

## D3: Winter dryness frequency

```{r D3}
show_question_section("D3")
```

## D4: Precipitation timing

```{r D4}
show_question_section("D4")
```

## D5: Dry-duration persistence

```{r D5}
show_question_section("D5")
```

# Fire

## F1: Overall frequency of high fire-weather days

```{r F1}
show_question_section("F1")
```

## F2: Summer frequency of high fire-weather days

```{r F2}
show_question_section("F2")
```

## F3: Fire-weather magnitude / intensity proxy

```{r F3}
show_question_section("F3")
```

## F4: Fire-season start timing

```{r F4}
show_question_section("F4")
```

## F5: Fire-season duration / persistence

```{r F5}
show_question_section("F5")
```

# Delta / ARIMA outputs

## What the Delta / ARIMA block is doing

For each metric and town-buffer combination:

1. Compute baseline anomalies relative to 1996-2025.
2. Compute **Sen's slope** for the observed series.
3. Fit a no-drift **ARIMA** model.
4. Simulate many synthetic series.
5. Compare the observed slope to the simulated distribution.
6. Assign a **Q1-Q7** category summarizing trend extremeness.

```{r arima-heatmap}
q_levels <- unique(question_outputs$question)
question_outputs$question <- factor(question_outputs$question, levels = q_levels)
question_outputs$qcat_num <- as.integer(factor(question_outputs$qcat, levels = paste0("Q", 1:7), ordered = TRUE))
question_outputs$unit <- paste(question_outputs$town, question_outputs$buffer, sep = " | ")

ggplot(question_outputs, aes(x = question, y = unit, fill = qcat_num)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_continuous(breaks = 1:7, limits = c(1, 7)) +
  labs(
    title = "ARIMA diagnostic category by question and location",
    x = "Question",
    y = "Town | Buffer",
    fill = "Qcat"
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_text(size = 7), panel.grid = element_blank())
```

```{r arima-qcat-counts}
question_outputs %>%
  count(question, qcat) %>%
  ggplot(aes(x = question, y = n, fill = qcat)) +
  geom_col() +
  labs(
    title = "Counts of ARIMA Qcat categories by question",
    x = "Question",
    y = "Count",
    fill = "Qcat"
  ) +
  theme_minimal(base_size = 11)
```

# PDI (Perception Δ Instrumentation)

This section summarizes the mismatch between survey perceptions and tenure-linked climate instrumentation.

```{r pdi-summary-table}
pdi_summary %>%
  arrange(question, .town) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(
    caption = "Town-level PDI summary by question",
    align = c("l", "l", rep("r", ncol(.) - 2))
  )
```

```{r pdi-mean-pdi-plot}
ggplot(pdi_summary, aes(x = .town, y = mean_pdi)) +
  geom_col() +
  facet_wrap(~ question, scales = "free_y") +
  labs(
    title = "Mean PDI by town and question",
    x = "Town",
    y = "Mean PDI"
  ) +
  theme_minimal(base_size = 11)
```

# Notes / interpretation guidance

- **Frequency metrics** generally increase when hot, dry, or high fire-weather days become more common.
- **Timing metrics** require directional interpretation: earlier onset often corresponds to smaller day values.
- **Magnitude metrics** summarize intensity or severity, not direct impacts such as burned area.
- **Qcat values** summarize how unusual an observed trend is relative to simulated ARIMA variability.
- **PDI values** indicate the degree of mismatch between reported perceptions and climate-derived instrumentation.
