---
title: "Idaho Heat, Drought, and Fire Metrics Report (By Question)"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
geometry:
  - margin=1in
header-includes:
  - \usepackage{graphicx}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.8,
  dpi = 300,
  fig.align = 'center',
  out.width = '100%'
)
```

# Overview

This report extends the original Idaho **fire metrics** workflow so it now covers the climate-comparable survey questions for **heat, drought, and fire** for the five Idaho towns:

- Deary
- Bovill
- Kendrick
- Juliaetta
- Elk River

The workflow still follows the same overall structure as your original fire report:

1. Read the Idaho town-level daily climate files.
2. Parse dates and water-year variables from `system.index`.
3. Build 1996–2025 daily climatologies and annual metrics.
4. Organize annual metrics by survey question.
5. Compute late-minus-early deltas and ARIMA-based trend categories.
6. Join to the survey responses and calculate tenure-based PDI summaries.

```{r source-pipeline, include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(knitr)

project_root <- getwd()
script_path <- file.path(project_root, "src", "extract_idaho_heat_drought_fire_metrics.R")
if (!file.exists(script_path)) {
  script_path <- "/mnt/data/extract_idaho_heat_drought_fire_metrics.R"
}
stopifnot(file.exists(script_path))
source(script_path)
```

# Data and metric construction

## Source data

This report uses the Idaho town/buffer daily climate files and the survey file `FW2L_DC.csv`.
Climate metrics are computed from **GridMET** only, restricted to:

- Deary
- Bovill
- Kendrick
- Juliaetta
- Elk River

Troy is excluded.

## Water-year conventions

- **Water year** runs from October 1 to September 30.
- `wyday` is the water-year day where October 1 = 1.
- Timing metrics for drought and fire therefore use **water-year timing**, not standard January-based Julian day.

## Climate-comparable survey questions included

```{r question-table}
question_labels %>%
  select(question, stressor, survey_col, question_text) %>%
  arrange(stressor, question) %>%
  kable(caption = "Question blocks included in the report")
```

# Annual metrics table preview

```{r annual-preview}
annual_all %>%
  arrange(town, buffer, wateryear) %>%
  head(15) %>%
  kable(caption = "Preview of annual metric table")
```

# Trend/PDI summary preview

```{r output-preview}
question_outputs %>%
  select(stressor, question, metric, town, buffer, sen_slope_per_decade, qcat, delta_late_minus_early) %>%
  arrange(stressor, question, metric, town, buffer) %>%
  head(20) %>%
  kable(caption = "Preview of question-level climate output summary")
```

```{r pdi-preview}
pdi_results$summary %>%
  arrange(question, .town) %>%
  kable(caption = "Preview of tenure-based PDI summary")
```

# Question-framed results

```{r helpers, include=FALSE}
plot_dir_local <- plot_dir
include_plot_if_exists <- function(path) {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat("(Plot not found)")
  }
}

question_order <- question_labels$question
```

```{r question-loop, results='asis'}
for (q in question_order) {
  q_info <- question_labels %>% filter(question == q)
  q_metrics <- q_metric_map %>% filter(question == q)
  q_out <- question_outputs %>% filter(question == q)
  q_pdi <- pdi_results$summary %>% filter(question == q)

  cat("\n\n## ", q, ": ", q_info$question_text[[1]], "\n\n", sep = "")
  cat("**Stressor:** ", q_info$stressor[[1]], "  \n", sep = "")
  cat("**Survey column:** `", q_info$survey_col[[1]], "`\n\n", sep = "")

  cat("### Metrics used\n\n")
  print(knitr::kable(q_metrics %>% select(metric), col.names = "Metric"))
  cat("\n\n")

  cat("### Climate summary by town/buffer\n\n")
  print(knitr::kable(
    q_out %>%
      select(town, buffer, metric, sen_slope_per_decade, qcat, delta_late_minus_early) %>%
      arrange(metric, town, buffer),
    digits = 3
  ))
  cat("\n\n")

  if (nrow(q_pdi) > 0) {
    cat("### Tenure-based PDI summary\n\n")
    print(knitr::kable(
      q_pdi %>%
        rename(town = .town) %>%
        select(town, n, mean_resp, mean_climate_delta_z, mean_pdi, sd_pdi),
      digits = 3
    ))
    cat("\n\n")
  }

  cat("### Example plots\n\n")
  for (m in q_metrics$metric) {
    ts_path <- file.path(plot_dir_local, paste0(q, "_timeseries_", m, ".png"))
    dist_path <- file.path(plot_dir_local, paste0(q, "_distribution_", m, ".png"))

    cat("**", m, "**\n\n", sep = "")
    include_plot_if_exists(ts_path)
    cat("\n\n")
    include_plot_if_exists(dist_path)
    cat("\n\n")
  }
}
```

# Output files written by the script

The sourced script writes the following files under `outputs/climate_questions/`:

- `idaho_annual_metrics_wide.csv`
- `idaho_annual_metrics_long.csv`
- `idaho_question_outputs.csv`
- `idaho_pdi_summary.csv`
- `idaho_pdi_person.csv`
- `idaho_pdi_deltas.csv`

It also writes per-metric plot PNGs under `plots/climate_questions/`.
